Low Infant Birth Weight in Brazil

### 2024-07-05 products/manuscript/Manuscript_Final.qmd
Draft 1 for advisor
Final changes for draft to advisor

```{r, echo=FALSE, message=FALSE}
#Loading packages
#install.packages("flextable")
#install.packages("gdtools")
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(gtsummary) #for summary tables
library(flextable) #for nice tables
```

# Abstract

This cross-sectional study uses linear regression models to understand disparities in infant birth weight in relation to maternal skin color (ancestry), age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s. Infants born to women of color (either Black or "*preta*" or mixed-race or "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant. I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, including intergenerational BLANK caused by racism, on population health in Brazil.

@@

At 4 in the morning on November 17, 1922, Virginia Maria da Conceição, a twenty-six-year-old Black ("*preta*") woman gave birth to a stillborn daughter in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now called Maternidade Escola). This was Conceição's fifth birth; prior, she had delivered three full-term infants vaginally (described as "naturally" in the clinical notes) and one premature infant, born at seven months gestational age. According to the notes, Conceição had had a "good" pregnancy. After arriving at the hospital, she remained in the first stage of labor ("dilatação") for five hours. The second stage of labor ("expulsão") lasted only thirty minutes. The placenta was delivered ten minutes later. The unnamed stillborn girl weighed 3,180 grams at birth, a normal weight for a full-term infant. Despite the apparently easy pregnancy and delivery, Conceição would not take home a healthy baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of the stillbirth [@rolindo1923, pp. 542-44].

Virginia Maria da Conceição's experience with stillbirth demonstrates the precarious state of maternal-infant health in early twentieth-century Brazil. I use Conceição's tragic but all too common story of poor perinatal and neonatal outcomes as a jumping off point to explore the historical relationship between social determinants of health and infant birth weight in a former slave society.

Public health and medical research over the past twenty years has shown that while race is a biological fiction, its social consequences have far-reaching influences on public health outcomes, particularly in countries with high levels of race-based inequality [@travassos2004]. Studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013; @wehby2015; @vilanova2019]. Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015; @vilanova2019]. Understanding these patterns is important as low birth weight is associated with higher rates of infant mortality in the first year of life [@cardoso2005; @vilanova2019].

These trends hold true in the United States, as well, where Black newborns have lower birth weights than White newborns, and these trends have worsened over time [@mangold1991; @pollock2021]. Racism is a major factor in these outcomes [@ifatunji2022], as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021; @clay2023; @mckenzie-sampson2021]. Other racial groups, whether U.S. or foreign-born, also face similar trends, although more research is needed to understand confounding factors [@montoya-williams2022].

Scholars across disciplines have argued that health disparities in countries with histories of race-based chattel slavery, from the United States to Brazil, are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @steckel1986]. In Brazil, most historical studies remain descriptive in nature and have not tested any quantitative associations between the legacies of slavery, including race-based inequality and its consequences and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2018a; @roth2018; @roth2020; @telles2022].

Here, I am to fill this gap by examining infant birth weight in relation to maternal skin color or ancestry, age, gestational status (parity), and nationality in Rio de Janeiro, Brazil's public maternity hospital, the Maternidade Laranjeiras, which opened in 1904 and provided free gynecological and obstetric care to the then-capital city’s poor and working-class women. It relies on a unique sample of infant birth weight data (n=2845).

@@

Brazil imported over 4 million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009]. It was also the last country to abolish slavery (1888) in the Western Hemisphere. It had, and still has, the largest number of African-descended peoples in the world outside of Nigeria [@michelle2023]. Around 55% of country's population today is of African-descent [@belandi2023].

Population health was poor for all people in the nineteenth century, but particularly so for enslaved individuals, who not only experienced epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies (@kiple1989).

In the immediate aftermath of the abolition of slavery, which coincided with the implementation of a republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into civil and political life occurred [@carvalho2004]. Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016; @roth2020].

Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019]. The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe. Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986; @benchimol1999; @hochman2013].  INFANT MORTALITY RATES IMPROVED. But the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s. Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].

By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to hospitalize labor and delivery [@barreto2015; @barreto2016; @brenes1991; @martins2004; @mott2002; @roth2020]. Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process. In hospitals, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine, including new cesarean section techniques [@martins2004; @rohden2009; @roth2021]. Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which harnessed advances in medical knowledge and public health infrastructure to support a technocratic model towards pregnancy, delivery, and motherhood. This ideology relied on essentialized notions of gender and race, foregrounding physicians’ scientific authority [@besse1996; @freire2009; @martins2008; @otovo2016].

Laranjeiras, which was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces. Maternidade Laranjeiras provided gynecological and obstetric care free of charge; thus, its clientele came from the city’s poor and working classes [@barreto2016; @roth2020]. The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants.

This study builds on an important body of work in economic history that uses hospital records to explore trends in birth weights in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998]. It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020].

## Data: Description and Sources

In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s first major obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s national medical association and obstetrics and gynecological society, both based in Rio de Janeiro. The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. [@roth2020; @roth2020a. For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository. The `Supplemental_Materials` includes information on data processing, exploratory analysis, and model fitting. All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available in the README. @roth2024.]

## Questions and Hypotheses

I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil measured using birth weight as a proxy for general fetal and neonatal health? I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro’s first public maternity hospital. I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), had lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.

To explore this relationship, I use linear models to estimate the relationship of maternal variables on infant birth weight. In the supplemental analysis, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable. Most results were insignificant, and thus I exclude them here. The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022]. In my main models, employ birthweight as a continuous variable in line with both current and historical studies [CITE], so these categories are not relevant for the linear models, although they can help explain the skew of the data. Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017]. I discuss this constraint in the limitations section below. Infant length was recorded in centimeters.

The covariates include maternal age (recorded as a continuous variable in years), maternal skin color, gestational status, and maternal nationality.

I ran two multivariate linear models: one in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent). In the second, I categorized maternal skin color into the dataset's original racial groups: White (the reference group), Black, and mixed race. Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011; @travassos2004]. Racial categories and skin color exist on a spectrum rather than a Black-White binary as in the United States [@telles2006]. Analyzing both specific categories of skin color and more general categories of ancestry is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015].

In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births. For the statistical analysis I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid any possible stillbirths [@nyarko2013; @wehby2015; @wehby2017].

## Descriptive analysis

I performed the exploratory analysis on the ML_summary dataset with n = 2845 observations and 14 variables. Table 1 demonstrates the distribution of the data.

```{r}
#| label: tbl-table1
#| tbl-cap: "Summary Statistics"
#| echo: FALSE
table1 <- readRDS(here::here("results", "tables", "table1_final.rds"))
flex_table <- as_flex_table(table1)
flex_table
```

ADD IN SECOND SUMMARY TABLE HERE

For maternal race (sample size n = 2695), the majority of the sample was White (*branca*) (42.45%). Mixed-race (*parda*) comprised 29.2%, and Black (*preta*) comprised 28.32%. If comparing White patients to all patients of color by combining Black and mixed-race categories, 1551 (57.6%) were of African descent (defined as *preta* or *parda*) and 1144 (42.4%) were of European descent (*branca*).

Of all reproductive outcomes (sample size n = 2761), 88% were natural deliveries. Yet, this also included spontaneous abortions. If we exclude spontaneous abortion (sample size n = 2672), 91% of outcomes were natural deliveries. The remaining 9% were interventionist or operative deliveries. These included cesarean sections or hysterectomies after a difficult delivery [@roth2020; @roth2020a]

For patients who went to the clinic to deliver their infant (thus excluding those suffering from spontaneous abortions or receiving postpartum care after an at-home delivery), 23 died. The Maternal Mortality Ratio (MMR), calculated as:

@@

where maternal deaths (MD) are divided by 10,000 live births (LB), was 94.26%. For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018; @roth2020]. But differential levels of recording must be taken into account. These rates could reflect a registration effect, as all births and deaths were recorded in the hospital, whereas accurate reporting at the city level was less reliable [@loudon1992]. Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s. Thus, the city’s rates were probably higher. Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures could be detrimental to maternal-infant health [@goldin1989].

@@

$$SBR = \frac{SB}{TB} \times 1,000$$ where the total number of stillbirths (SB) is divided by 1000 total births (TB). The SBR is 84.8% in this specific clinic. We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% [@roth2020; @roth2018]. Again, registration effects could explain the higher rates in the hospital [@loudon1992].

@@

where the total number of live male births (M) is divided by by 100 live female births (F) in a given period.

The sex ratio is 1.2: there were 120 male live births per 100 female live births at this specific clinic. This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024]. Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female. Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016]. However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants. The skewed sex ratio deserves further study.

@@

I ran four linear models to understand the relationship between maternal factors and infant birth weight. The first two models are a simple and multiple linear regression, respectively, with outcome variable as birth weight in grams. In these two models, I recoded maternal skin color as maternal ancestry (`Euro-descent` or `Afro-descent`).

The first simple linear regression model is as follows:

@@

The second multiple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 3.

@@

The second two models are also simple and multiple linear regressions with outcome variable as birth weight in grams. However, in these two models, I maintained maternal skin color as the original three categories recorded in the hospital records, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women. The first simple linear regression model is as follows:

@@

The second multiple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term. Results are displayed in Table 5.

@@

## Results

All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference group) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). Age and nationality were not significant. Women who were nulliparous (delivering their first child) had infants with lower birth weights.

paradox between taller people and racial disparities in infant birth weight 
catch up if surviving childhood (Steckel)

# Discussion

Syphilis
prenatal care had an 8.22% positive rate for prenatal syphilis in 1935 and an 8.66% positive rate in 1936 (Fontenelle 326).

Between 1935 and 1938 7.92% positive tests (Fontenelle, 275 second Fontenelle).

Racism
Malnutrition during pregnancy (not linear)

## Strengths and Limitations

>>>>

```{r, echo=FALSE, message=FALSE}
#Loading packages
#install.packages("flextable")
#install.packages("gdtools")
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(gtsummary) #for summary tables
library(flextable) #for nice tables
```
# Abstract

This cross-sectional study uses linear models to understand disparities in infant birth weight in relation to maternal skin color, age, parity, and nationality in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now Maternidade Escola) in the 1920s. It finds that infants born to women of color (either Black, "*preta,*" or mixed-race, "*parda*") had lower birth weights than infants born to White women, whether Brazilian or immigrant. I hypothesize these differences were related to the historical legacies of slavery, only abolished in 1888, including race-based socioeconomic inequality and possible epigenetic effects from the lived experience of racism, on population health in Brazil.

@@

At four in the morning on November 17, 1922, Virginia Maria da Conceição, a twenty-six-year-old Black ("*preta*") woman gave birth to a stillborn daughter in Rio de Janeiro, Brazil's first public maternity hospital, the Maternidade Laranjeiras (now called Maternidade Escola). This was Conceição's fifth birth; prior, she had delivered three full-term infants vaginally and one premature infant, born at seven months gestational age. According to the hospital records, Conceição had had a "good" pregnancy, with a relatively easy delivery. After arriving, she remained in the first stage of labor ("dilatação") for five hours. The second stage of labor ("expulsão") lasted thirty minutes. She delivered the placenta ten minutes later. The unnamed stillborn girl weighed 3180 grams at birth, a normal weight for a full-term infant and around 150 grams higher than the average weight for female infants and for Black infants born at the hospital in the 1920s. But, as the hospital records detailed, Conceição did not take home a healthy baby; rather, "hereditary syphilis," as the fetal autopsy noted, was the cause of intrauterine stillbirth [@rolindo1923, pp. 542-44].

Virginia Maria da Conceição's experience with syphilis and stillbirth demonstrates the precarious state of maternal-infant health in early twentieth-century Brazil. I use Conceição's tragic but all too common story of poor perinatal and neonatal outcomes as a jumping off point to explore the historical legacy of slavery on maternal-infant health in Brazil.

Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. Contemporary studies in Brazil have shown that non-White mothers give birth to infants at lower birth weights than their White counterparts [@nyarko2013; @wehby2015a; @vilanova2019]. Mothers with lower education levels and lower socioeconomic status also give birth to infants at lower birth weights [@wehby2015a; @wehby2017a; @vilanova2019].

These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns have lower birth weights than White newborns, with trends worsening over time [@pollock2021]. Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have infants with lower birth weights and higher rates of pre-term birth than Black immigrant women [@braveman2021a; @clay2023a; @ifatunji2022a; @mckenzie-sampson2021]. Other racial groups, whether U.S. or foreign-born, also face similar trends, although more research is needed to understand confounding factors [@montoya-williams2022]. Understanding these patterns within and across communities is important as low birth weight is associated with higher rates of infant mortality in the first year of life and possible health problems over the life course [@cardoso2005; @vilanova2019].

Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery, including the United States, are longstanding and tied to the unequal and violent social relations produced under that institution [@hogarth2019; @jasienska2009; @owens2019; @steckel1986]. In Brazil, most historical studies of health under slavery and in its aftermath remain descriptive in nature and have not tested any quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@otovo2016; @roth2018a; @roth2018; @roth2020; @telles2022].

Here, I aim to fill this gap by examining infant birth weight in relation to maternal skin color, age, parity, and nationality in the Maternidade Laranjeiras, which opened in 1904 and provided free gynecological and obstetric care to the then-capital city’s poor and working-class women.

In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon. In our current postgenomic era, in which "race" as a biological category has been thoroughly debunked, the possibility for centuries-old patterns of hierarchical scientific thinking remains very real [@alam2024; @desalle2018; @roberts2011]. Discussions of the potential effects of racism on epigenetic markers is one area where this essentialization can occur. Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022]. Poor Black women like Virginia Maria da Conceição and her children were not "pathologically" different or part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].

This paper aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine the lives of people like Virginia Maria da Conceição.

@@

Brazil imported over four million enslaved Africans during the nearly four centuries of the existence of chattel slavery in the country [@klein2009]. It was also the last country to abolish slavery (1888) in the Western Hemisphere. In the nineteenth century, population health was poor for all people, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from severe nutritional deficiencies [@karasch1987; @kiple1989].

In the immediate aftermath of the abolition of slavery, which coincided with the implementation of the country's first republican form of government in 1889, no state-run efforts to incorporate formerly enslaved, African-descended peoples into civil and political life occurred [@carvalho2004]. Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower educational and literacy levels, and had worse health outcomes including infant mortality than their White and White immigrant counterparts [@ball2020; @fischer2008; @otovo2016; @roth2020].

Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019]. The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe. Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@anderson1986; @benchimol1999; @hochman2013]. Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020]. Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].

By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2015; @barreto2016; @brenes1991; @martins2004; @mott2002; @roth2020]. Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process. There, obstetricians and gynecologists began implementing surgical advances in the realm of women’s medicine [@martins2004; @rohden2009; @roth2021]. Underlying these structural and technological changes was the scientific motherhood movement, supported by both physicians and elite women, which harnessed advances in medical knowledge and public health infrastructure to support a technocratic model towards pregnancy and motherhood. This ideology relied on essentialized notions of gender and race, foregrounding physicians’ scientific authority [@besse1996; @freire2009; @martins2008; @otovo2016].

At the same time, old racial hierarchies based on legal status morphed into new "scientific" ways to categorize humans based on skin color. In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than coercively restricting reproduction through sterilization [@schwarcz1993; @stepan1991]. Although some eugenicists called for sterilization of "unfit" citizens, they never couched these discussions in explicitly racial terms [@roth2020b; @wegner2013]. In the end, however, whiteness still remained a cultural and social "goal" even if the state did not engage in coercive measures of reproductive control.

Maternidade Laranjeiras, which was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces. Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020]. The majority of patients were women of color defined as either Black or mixed race, and the majority of White patients were immigrants. Because most clinical notes were short – only those that described surgical births such as cesarean sections, or in the case of Virginia Maria da Conceição included the fetal autopsy report, had longer notes – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021]. Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.

This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998]. It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].

## Analysis

I analyzed a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras.[^1]

[^1]: For complete information on the source, see @roth2020; @roth2020a. My Github repository, in particular, the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting. All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online. See @roth2024.

Relying on the dataset, I ask whether the legacy of slavery, both as possible intergenerational inheritance patterns or as contemporary experiences of discrimination in a post-abolition society, affected the health of infants born in the first public maternity hospital in 1920s Brazil using birth weight as a proxy for general fetal and neonatal health. I quantify how maternal race, nationality, age, and parity explain racial disparities in infant birth weight.

```{r}
#| label: tbl-table1
#| tbl-cap: "Summary Statistics"
#| echo: FALSE
table1 <- readRDS(here::here("results", "tables", "table1_final.rds"))
flex_table <- as_flex_table(table1)
flex_table
```
{{< pagebreak >}}

Table 1 provides summary statistics for the sample. When comparing White patients to all patients of color by combining Black and mixed-race categories, 1551 (58%) were of African descent (defined as *preta* or *parda*) and 1144 (42%) were of European descent (*branca*). First-time mothers ("nulliparous") were 41% of the sample (n = 2836). Perhaps they would return to Laranjeiras, or another maternity hospital, to give birth to their future children, demonstrating the process of hospitalization of childbirth in real time. Most mothers were Brazilian. Of the total sample (n = 2773), 84% were Brazilian. Most immigrants were European.

Of all reproductive outcomes (n = 2761), 88% were natural deliveries. If we exclude spontaneous abortion from the sample (n = 2672), 91% of outcomes were natural deliveries. The remaining 9% were interventionist or operative deliveries. These included forceps and cesarean sections [@roth2020; @roth2020a].

For patients who went to the clinic to deliver their infant (thus excluding those suffering from spontaneous abortions), 23 died. The Maternal Mortality Ratio (MMR), calculated as:

@@

where maternal deaths (MD) are divided by 10,000 live births (LB), was 94.26%. For those same years, the MMR for the city of Rio de Janeiro was 65.65% [@roth2018; @roth2020]. This is a stark difference, but we must take into account differential levels of recording. Hospital rates could have reflected a registration effect, as all births and deaths were recorded in the hospital, whereas reporting at the city level was less reliable [@loudon1992]. Rio de Janeiro’s vital statistics were still poorly defined and intermittently collected in the 1920s [@roth2020]. Thus, the city’s rates were probably higher. Nonetheless, the difference demonstrates that delivering in the presence of licensed clinicians did not necessarily improve outcomes for the mother and supports other historical studies that interventionist obstetric procedures could be detrimental to maternal-infant health before the rise of antibiotics and blood transfusions [@goldin1989].

@@

$$SBR = \frac{SB}{TB} \times 1,000$$ 
where the total number of stillbirths (SB) is divided by 1000 total births (TB), was 84.8%. We can compare this to the mean SBR for the city of Rio de Janeiro for the same period, which was 73.68% [@roth2020; @roth2018]. Again, registration effects could explain the higher hospital rates.

@@

where the total number of live male births (M) is divided by 100 live female births (F) in a given period.

The sex ratio was 1.2: there were 120 male live births per 100 female live births at this specific clinic. This is much higher than the current range of between 103 and 107 male births per 100 female births [@ritchie2024]. Historically, a skewed sex ratio suggests that preferential infanticide or abortion was occurring -- parents were more likely to terminate a pregnancy or kill an infant if it was a female. Contrary to popular belief, this practice occurred in both Asian and European countries, although no evidence of it exists for the Americas [@hanlon2016]. However, this explanation does not hold for a maternity clinic in which women were seeking care to deliver their infants. The skewed sex ratio deserves further study.

Table 2 provides the mean birth weights for the Maternidade Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand.

```{r}
#| label: tbl-table2
#| tbl-cap: "Mean Birth Weights"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
table2
```

As we can see, Black infants born at the Maternidade Laranjeiras had the lowest mean birth weight (3037 grams) compared to all other historical samples except for estimated enslaved birth weights in the antebellum US South, which scholars have since called into question for being too low [@costa2004]. Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent time period. Yet, they were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. Yet they were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent time period (3423 grams). As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same time period [@costa1998; @costa2004; @goldin1989].

@@

I use Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight. In the supplemental analysis available online, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable (either normal birth weight, NBW, or low birth weight, LBW, according to current WHO classifications), and in line with some studies on birth weight [@worldhealthorganization2022; @falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a]. Most logistic model results were insignificant, and thus I exclude them here. In my main models, I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is best practice for disentangling pre-term birth (PTB) from low birth weight (LBW) [@cutland2017]. I discuss this constraint in the discussion below. The covariates include maternal age (recorded as a continuous variable in years), maternal skin color, gestational status, and maternal nationality.

I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race. Racial terminology in Brazil was, and continues to be, complex and dynamic [@alberto2011; @travassos2004]. Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006]. Analyzing both specific categories of skin color and more general categories of ancestry can help address these different understandings of race and is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

Yet, neither the records' reporting of maternal skin color nor my own reclassification of the skin color of the mothers in this study can fully serve as a proxy for maternal health. Assuming so runs the potential of reifying intra-group differences at the expense of inter-group ones. Yet, given the limited historical data available, I run these models understanding the inherent limitations of using skin color as a category at all and as a proxy for health status in particular.

The simple linear regression model is as follows:

@@

The multiple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 3.

@@

The second two models are also simple and multiple linear regressions. However, in these two models, I maintained maternal skin color as the original three categories recorded in the hospital records, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women. The simple linear regression model is as follows:

@@

The multiple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term. Results are displayed in Table 4.

@@

All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference group) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). Age and nationality were not significant. Women who were nulliparous (delivering their first child) had infants with lower birth weights, which is an association still found today [@souza2020].

# Discussion and Conclusion

The intrauterine environment is a critical determinant of infant birth weight. Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care can all influence birth weight. Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother [@costa1998; @steckel1986]. A threshold of malnutrition must be met before there is an adverse effect on fetal growth. Other factors include maternal age, parity, and birth spacing, as well as gestational age. We can imagine that poorer women performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992]. Generalized poverty might have influenced maternal nutrition, as well as the ability to access prenatal care.
